Cultural Transmission Can Explain the Evolution of Cooperation

### 2020-07-30 ms/main.tex
format result 3; edit appendix b

\begin{result}[Vertical, oblique, and horizontal transmission of cooperation]
The cultural evolution of a cooperator phenotype will follow one of the following scenarios, depending on the horizontal transmission bias $T_A-T_B$ and the coefficients $\beta_1$ and $\beta_3$:
\begin{enumerate}
\item \emph{Fixation of the cooperative phenotype $A$}, 
\begin{enumerate}
\item if $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$, or
\item if $T_A>T_B$ and $0<\beta_3$, or 
\item if $T_A<T_B$ and $\beta_1<\beta_3$.
\end{enumerate}

\item \emph{Fixation of the defector phenotype $B$}, 
\begin{enumerate}
\item if $T=T_A=T_B$ and $c > b\cdot \frac{\alpha T}{1-T}$, or 
\item if $T_A>T_B$ and $\beta_1<\beta_3<0$, or 
\item if $T_A<T_B$ and $\beta_3<0$.
\end{enumerate}

\item \emph{Protected polymorphism, or co-existence of both phenotypes}, if $T_A < T_B$ and $0<\beta_3<\beta_1$.

\item \emph{Fixation of either phenotype depending on initial frequency}, if $T_A>T_B$ and $\beta_3<\beta_1$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany and Ayelet Shavit for discussions and comments.
This work was supported in part by
the Israel Science Foundation 552/19 (YR),
and Minerva Stiftung Center for Lab Evolution (YR).
% TODO funding for other authors?
}

@@

We show that $f'(\hat{p}^*)<0$ is a sufficient condition for local stability an equilibrium $\hat{p}^*$.
We will write $f(\hat{p})$ as a Taylor approximation around the equilibrium $\hat{p}^*$. 
\begin{equation} \label{eq:appendix_b_taylor}
  f(\hat{p}) = \bar{w}(\hat{p}'-\hat{p}) = f(\hat{p}^*) + f'(\hat{p}^*)(\hat{p}-\hat{p}^*) + R_2(\hat{p})
\end{equation}
Where $R_2(\hat{p})$ is the remainder. Since $\hat{p}^*$ is an equilibrium $f(\hat{p}^*)=0$.
\begin{equation} \label{eq:appendix_b_taylor0}
  f(\hat{p}) = \bar{w}(\hat{p}'-\hat{p}) =  f'(\hat{p}^*)(\hat{p}-\hat{p}^*) + R_2(\hat{p})
\end{equation}
Neglecting the remainder(?) and we get: 
\begin{equation} \label{eq:appendix_b_taylor_without_remainder}
  f(\hat{p}) = \bar{w}(\hat{p}'-\hat{p}) = f'(\hat{p}^*)(\hat{p}-\hat{p}^*)
\end{equation}
And we get:
\begin{equation} \label{eq:appendix_b_derivative}
  \begin{aligned}
  f'(\hat{p}^*) &= \bar{w}\Big(\frac{\hat{p}'-\hat{p}}{\hat{p}-\hat{p}^*}\Big)\\
  & = \bar{w}\Big(\frac{\hat{p}'-\hat{p}^*+\hat{p}^*-\hat{p}}{\hat{p}-\hat{p}}\Big) 
  = \bar{w}\Big(\frac{\hat{p}'-\hat{p}^*}{\hat{p}-\hat{p}^*}-1\Big) .
  \end{aligned}
\end{equation}
In order for the equilibrium $\hat{p}^*$ to be stable we demand that if $\hat{p} > \hat{p}^*$ then $\hat{p}'-\hat{p}^* < \hat{p}-\hat{p}^*$
and if $\hat{p} < \hat{p}^*$ we demand that $-(\hat{p}'-\hat{p}^*) < -(\hat{p}-\hat{p}^*)$. 
In other words, $\frac{\hat{p}'-\hat{p}^*}{\hat{p}-\hat{p}^*}<1$ for every $\hat{p}$.
Therefore,
\begin{equation} \label{eq:appendix_b_condition}
  \begin{aligned}
  f'(\hat{p}^*) = \bar{w}\Big(\frac{\hat{p}'-\hat{p}^*}{\hat{p}-\hat{p}^*}-1\Big) < 0 .
  \end{aligned}
\end{equation}
Therefore, $f'(\hat{p}^*)$ is a sufficient condition for stability of $\hat{p}^*$.

>>>>

\begin{result}[Vertical, oblique, and horizontal transmission of cooperation]
The cultural evolution of a cooperator phenotype will follow one of the following scenarios, depending on the horizontal transmission bias $T_A-T_B$ and the coefficients $\beta_1$ and $\beta_3$:
\begin{enumerate}
\item \emph{Fixation of cooperation}, 
if $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$; or
if $T_A>T_B$ and $0<\beta_3$; or 
if $T_A<T_B$ and $\beta_1<\beta_3$.
\item \emph{Fixation of the defection}, 
if $T=T_A=T_B$ and $c > b\cdot \frac{\alpha T}{1-T}$; or 
if $T_A>T_B$ and $\beta_1<\beta_3<0$; or 
if $T_A<T_B$ and $\beta_3<0$.
\item \emph{Coexistence of both phenotypes at $\hat{p}^*$},
if $T_A < T_B$ and $0<\beta_3<\beta_1$.
\item \emph{Fixation of either phenotype depending on initial frequency}, if $T_A>T_B$ and $\beta_3<\beta_1$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany and Ayelet Shavit for discussions and comments.
This work was supported in part by
the Israel Science Foundation 552/19 (YR),
and Minerva Stiftung Center for Lab Evolution (YR).
% TODO funding for other authors?
}

@@

Denote $f(p)=\lambda(p'-p)$, where $\lambda>0$, and assume $f(p^*)=0$.
We want a condition for $|p'-p^*| < |p-p^*|$.

If $p>p^*=0$, we want a condition for $p'<p$, or
$\frac{p'}{p}<1$, or
$\lambda \frac{p'-p}{p} < 0$, or
$\frac{f(p)}{p} < 0$.
Using a linear approximation for $f(p)$ near $0$, we have
\begin{equation}\begin{aligned}
p' &< p \Leftrightarrow \\
\frac{f'(0) \cdot p + O(p^2)}{p} &< 0 \Leftrightarrow \\
f'(0) + O(p) &< 0.
\end{aligned}\end{equation}
Therefore, by definition of big-O notation, if $f'(0)<0$ then there exists $\epsilon>0$ such that for any $0<p<\epsilon$ it is guaranteed that $0<p'<p$, that is, $p'$ is closer than $p$ to zero.

If $p<p^*=1$, we want a condition for $1-p' < 1-p$, or
$\frac{1-p'}{1-p}<1$, or
$\lambda \frac{-(p'-p)}{1-p} < 0$, or
$-\frac{f(p)}{1-p} < 0$.
Using a linear approximation for $f(p)$ near $1$, we have
\begin{equation}\begin{aligned}
1-p' &< 1-p  \Leftrightarrow \\
\frac{f'(1)(p-1) + O\big((p-1)^2\big)}{p-1} &< 0 \Leftrightarrow \\
f'(1) - O(1-p) &< 0.
\end{aligned}\end{equation}
Therefore, if $f'(1)<0$ then there exists $\epsilon>0$ such that for any $1-\epsilon<1-p<1$ it is guaranteed that $1-p'<1-p$, that is, $p'$ is closer than $p$ to one.
